---
title: "BPS"
format: html
editor: visual
---

## Loading libraries

```{r}
library(tidyverse)
library(readxl)
library(rriskDistributions)
library(writexl)
```

## Calculating the Danish baseline for sperm motility

```{r}
getwd()

data_app <- read_csv("/Users/idasofieadrian/Desktop/DTU/SPECIALE/Speciale/data/DATAapplicants.csv")
data_donor <- read_csv("/Users/idasofieadrian/Desktop/DTU/SPECIALE/Speciale/data/DATAdonors.csv")

#data_donor_first <- data_donor %>%
  #arrange(DonorID, EjacYR, EjacMO, EjacDAY) %>%  # ensures chronological order within donor
  #group_by(DonorID) %>%
  #slice(1) %>%                                   # keeping only first donation
  #ungroup()

data_donor_best <- data_donor %>%
  group_by(DonorID) %>%
  slice_max(TMSconc, n = 1, with_ties = FALSE) %>%
  ungroup()

# Function to calculate 32% motility
compute_p0 <- function(df, thr = 32) {
  df2 <- df %>%
    mutate(
      # calculating the percentage of sperms cells that are progressively motile
      prog_motility_pct = 100 * (TMSconc) / TSconc
    ) %>%
    filter(
      # handling impossible or invalid values, fx TSconc = 0, division would give inF or NaN
      is.finite(prog_motility_pct),
      prog_motility_pct >= 0,
      prog_motility_pct <= 100
    )
  
  list(
    p0 = mean(df2$prog_motility_pct < thr), # mean treats TRUE/FALSE values as numbers
    n = nrow(df2)
  )
}



# use function on applicant data
print("applicant data")
compute_p0(data_app)

# here also on donor data
print("donor data")
compute_p0(data_donor_best)


# saving applicant motitlity data just in case
df_mot <- compute_p0(data_app)

p0_motility_dk <- df_mot$p0


# COULD THIS BE USED FOR SOMETHING? NOT UTILIZED RIGHT NOW

# why do many donors also not meet the 32% threshold criterion? Only the first sample is utilized, maybe that sample is not others samples of the donor are better? Maybe they fullfill other criteria?

# checked by selecting best concentration, much more applicable results :-)

# creating a new dataframe, with a column describing the motility percentage for each applicant
data_app_mot <- data_app %>%
  mutate(
    prog_motility_pct = 100 * TMSconc / TSconc
  ) %>%
  filter(
    is.finite(prog_motility_pct),
    prog_motility_pct >= 0,
    prog_motility_pct <= 100
  )

```

## Importing BPS levels for Danish men

```{r}
#getwd()

# assumes log normal distribution
# sperm attributes are always positive

df <- read_excel("/Users/idasofieadrian/Desktop/DTU/SPECIALE/Speciale/data/MEN BPS.xlsx") %>%
  mutate(P25 = as.numeric(P25)) %>%
  filter(Year == 2017) %>%
  mutate(
    p0_motility_dk     = p0_motility_dk, # added but not used right now
    meanlog_dk         = log(P50),
    sdlog_dk           = (log(P75) - log(P25)) / (qnorm(0.75) - qnorm(0.25)),
    p_detectable_dk    = 1 - plnorm(LOD, meanlog_dk, sdlog_dk),
    mean_exp_dk        = exp(meanlog_dk + 0.5 * sdlog_dk^2),  # arithmetic mean of full distribution on log scale
    median_dk          = P50  
  )
```

## Chinese ORs

```{r}
# ORs from the Chinese paper
# Table 3
OR_q1 <- 1.00
OR_q2 <- 0.94
OR_q3 <- 1.04
OR_q4 <- 1.62

# Convert 95% CI to log-scale SE
# simulating uncertainty
LCL_q2 <- 0.62; UCL_q2 <- 1.44
LCL_q3 <- 0.68; UCL_q3 <- 1.58
LCL_q4 <- 1.07; UCL_q4 <- 2.43

se_q2 <- (log(UCL_q2) - log(LCL_q2)) / (2 * 1.96)
se_q3 <- (log(UCL_q3) - log(LCL_q3)) / (2 * 1.96)
se_q4 <- (log(UCL_q4) - log(LCL_q4)) / (2 * 1.96)
```

## Distribution

```{r}
# from Table 2
# Using p25, p50, and p75 to fit a lognormal distribution
# creatine adjusted and for two samples
fit_chi <- get.lnorm.par(
  p = c(0.25, 0.50, 0.75),
  q = c(0.13, 0.29, 0.69),
  show.output = FALSE,
  plot = TRUE
)


# extract the fitted lognormal parameters as ordinary numbers
meanlog_chi <- unname(fit_chi["meanlog"])
sdlog_chi   <- unname(fit_chi["sdlog"])


# Simulate Chinese detectable exposure distribution
set.seed(123)
x_chi <- rlnorm(1e6, meanlog = meanlog_chi, sdlog = sdlog_chi)

# Quartile cutpoints among Chinese BPS values
cut1 <- qlnorm(0.25, meanlog = meanlog_chi, sdlog = sdlog_chi)
cut2 <- qlnorm(0.50, meanlog = meanlog_chi, sdlog = sdlog_chi)
cut3 <- qlnorm(0.75, meanlog = meanlog_chi, sdlog = sdlog_chi)

# Mean dose within each quartile
dose_q1 <- mean(x_chi[x_chi <= cut1])
dose_q2 <- mean(x_chi[x_chi > cut1 & x_chi <= cut2])
dose_q3 <- mean(x_chi[x_chi > cut2 & x_chi <= cut3])
dose_q4 <- mean(x_chi[x_chi > cut3])


## Build dose-response dataset and fit linear trend ##
# dose = representative exposures (means) for the 4 quartiles
# dose_c = making tertile 1 the reference point/baseline where log(RR)=0, RR=1
# logrr = converting the RR to the log scale
# se = using the SE for the log-risk estimates, NA for q1 since it is the reference
study_dat <- tibble(
  dose   = c(dose_q1, dose_q2, dose_q3, dose_q4),
  dose_c = c(dose_q1, dose_q2, dose_q3, dose_q4) - dose_q1,
  logrr  = log(c(OR_q1, OR_q2, OR_q3, OR_q4)),
  se     = c(NA, se_q2, se_q3, se_q4)
)

# Derive CI bounds from SE for plotting
study_dat <- study_dat %>%
  mutate(
    logrr_lower = logrr - 1.96 * se,
    logrr_upper = logrr + 1.96 * se
  )


# fitting linear model, trying to finde the slope beta of the best straight line
# lm(logrr ~ 0 + dose_c) fits the line without an intercept to force it through (0,0) 
# this means dose_c = 0, logrr = 0 which makes 1 at reference dose (RR = e^logrr = 1, where logrr = 0)
# data = study_dat[2:3, ], uses just tertile 2 and 3 (removes the first row) because tertile 1 is reference

fit_lin <- lm(
  logrr ~ 0 + dose_c,
  data = study_dat[2:4, ],
  weights = 1 / study_dat$se[2:4]^2
)

beta_hat <- unname(coef(fit_lin)["dose_c"])
beta_se  <- summary(fit_lin)$coefficients["dose_c", "Std. Error"]

# setting q1 as the reference
x_ref <- dose_q1

beta_hat
beta_se
summary(fit_lin)
```

```{r}
# making a sequence of dose values for plotting
# Model uncertainty added
pred_grid <- tibble(
  dose = seq(min(study_dat$dose), max(study_dat$dose), length.out = 100)
) %>%
  mutate(dose_c = dose - dose_q1) %>%
  bind_cols(
    predict(fit_lin, newdata = ., interval = "confidence", level = 0.95) %>%
      as_tibble() %>%
      rename(logrr_fit = fit, logrr_lwr = lwr, logrr_upr = upr)
  )

# plot
p_logrr <- ggplot() +
  # Model confidence band
  geom_ribbon(
    data = pred_grid,
    aes(x = dose, ymin = logrr_lwr, ymax = logrr_upr),
    fill = "#045A8DFF",
    alpha = 0.15
  ) +
  # Model fit line
  geom_line(
    data = pred_grid,
    aes(x = dose, y = logrr_fit),
    color = "#045A8DFF",
    linewidth = 1
  ) +
  # Data points with error bars  (requires logrr_lower / logrr_upper in study_dat)
  geom_errorbar(
    data = study_dat,
    aes(x = dose, ymin = logrr_lower, ymax = logrr_upper),
    color = "#045A8DFF",
    width = 0.02,
    linewidth = 0.6
  ) +
  geom_point(
    data = study_dat,
    aes(x = dose, y = logrr),
    color = "#045A8DFF",
    size = 3
  ) +
  labs(
    title = "BPS and motility risk",
    x = "Urinary BPS concentration",
    y = "log(OR)"
  ) +
  theme_classic() +
  theme(plot.title = element_text(size = 16, face = "bold"))

p_logrr

# saving plot
ggsave(
  filename = "logrr_linear_model_bps_semen.png",
  plot = p_logrr,
  width = 7,
  height = 5,
  dpi = 300
)




```

## YLD inputs

```{r}
## GBD inputs: male infertility prevalence (Denmark, 15-49, 2023)

prevalence_mode  <- 12120.59
prevalence_lower <- 6434.65
prevalence_upper <- 21390.85

## Disability weight (GBD, male infertility)

DW_mode <- 0.007564045
DW_min  <- 0.003335558
DW_max  <- 0.015242128

# LOW DW, NOT LIFE THREATENING


### YLD FROM GBD ###
# 78.41 (25.71, 180.01)

dw_test <- 78.41/12120.59

# NO YLL SINCE INFERTILITY DOES NOT CAUSE DEATH
# YLD = AC * DW * D
# NO DURATION IN THIS CASE
```

## PERT

```{r}
## Helper functions

rpert <- function(n, min, mode, max) {
  alpha <- 1 + 4 * (mode - min) / (max - min)
  beta  <- 1 + 4 * (max - mode) / (max - min)
  min + rbeta(n, alpha, beta) * (max - min)
}
```

## Monte Carlo

```{r}
run_mc_row_semen <- function(
  p_detectable_dk, median_dk,
  prevalence_mode, prevalence_lower, prevalence_upper,
  DW_min, DW_mode, DW_max,
  beta_hat, beta_se, x_ref
) {

  # uncertainty in dose-response slope
  beta <- rnorm(1, mean = beta_hat, sd = beta_se)

  # RR for poor motility at mean detectable Danish exposure
  # log(RR) = beta * (dose - reference dose)
  rr_motility <- exp(beta * (median_dk - x_ref))

  # PAF for poor motility
  PAF_motility <- (p_detectable_dk * (rr_motility - 1)) /
                  (1 + p_detectable_dk * (rr_motility - 1))

  # mediation fraction: proportion of male infertility driven by poor motility
  # CONSERVATIVE APPROACH
  # med_frac <- rpert(1, min = 0.16, mode = 0.20, max = 0.24)
  # Moderate approach
  med_frac <- rpert(1, min = 0.19, mode = 0.40, max = 0.60)

  # PAF for male infertility
  # infertility cases caused by motility
  PAF_infertility <- PAF_motility * med_frac

  # prevalence uncertainty
  # does this align with the correct approach?
  prev <- rpert(1,
                min  = prevalence_lower,
                mode = prevalence_mode,
                max  = prevalence_upper)

  # attributable cases
  AC <- PAF_infertility * prev

  # disability weight uncertainty
  DW <- rpert(1, DW_min, DW_mode, DW_max)

  # YLD (prevalence-based, no duration multiplier)
  YLD  <- AC * DW
  DALY <- YLD   # no YLL for male infertility

  tibble(
    PAF_motility    = PAF_motility,
    PAF_infertility = PAF_infertility,
    med_frac        = med_frac,
    AC              = AC,
    DW              = DW,
    YLD             = YLD,
    DALY            = DALY
  )
}

run_mc_summary_semen <- function(
  p_detectable_dk, median_dk,
  prevalence_mode, prevalence_lower, prevalence_upper,
  DW_min, DW_mode, DW_max,
  beta_hat, beta_se, x_ref,
  nsim = 3000
) {

  sims <- map_dfr(seq_len(nsim), ~ run_mc_row_semen(
    p_detectable_dk  = p_detectable_dk,
    median_dk        = median_dk,
    prevalence_mode  = prevalence_mode,
    prevalence_lower = prevalence_lower,
    prevalence_upper = prevalence_upper,
    DW_min           = DW_min,
    DW_mode          = DW_mode,
    DW_max           = DW_max,
    beta_hat         = beta_hat,
    beta_se          = beta_se,
    x_ref            = x_ref
  ))

  tibble(
    PAF_motility_med    = median(sims$PAF_motility),
    PAF_motility_lo     = quantile(sims$PAF_motility, 0.025),
    PAF_motility_hi     = quantile(sims$PAF_motility, 0.975),

    PAF_infertility_med = median(sims$PAF_infertility),
    PAF_infertility_lo  = quantile(sims$PAF_infertility, 0.025),
    PAF_infertility_hi  = quantile(sims$PAF_infertility, 0.975),

    AC_med  = median(sims$AC),
    AC_lo   = quantile(sims$AC, 0.025),
    AC_hi   = quantile(sims$AC, 0.975),

    YLD_med  = median(sims$YLD),
    YLD_lo   = quantile(sims$YLD, 0.025),
    YLD_hi   = quantile(sims$YLD, 0.975),

    DALY_med = median(sims$DALY),
    DALY_lo  = quantile(sims$DALY, 0.025),
    DALY_hi  = quantile(sims$DALY, 0.975)
  )
}
```

```{r}
results_semen <- df %>%
  mutate(
    mc = pmap(
      list(p_detectable_dk, median_dk),
      ~ run_mc_summary_semen(
        p_detectable_dk  = ..1,
        median_dk        = ..2,
        prevalence_mode  = prevalence_mode,
        prevalence_lower = prevalence_lower,
        prevalence_upper = prevalence_upper,
        DW_min           = DW_min,
        DW_mode          = DW_mode,
        DW_max           = DW_max,
        beta_hat         = beta_hat,
        beta_se          = beta_se,
        x_ref            = x_ref,
        nsim             = 3000
      )
    )
  ) %>%
  unnest(mc)

results_semen
```

## checks

```{r}

df %>%
  mutate(rr_at_median = exp(beta_hat * (median_dk - x_ref))) %>%
  dplyr::select(Year, median_dk, p_detectable_dk, rr_at_median)

beta_hat
x_ref


dose_q1
dose_q2
dose_q3
dose_q4
```

## Write to excel

```{r}
#| eval: false
#| include: false
getwd()
#write_xlsx(results_semen, "results_semen_china.xlsx")
```

# BOTTOM UP APPROACH

```{r}
# reported percentages of males falling below the 32% threshold
p0_motility_china <- 0.275

# Danish men falling below threshold
p0_motility_dk

# counterfactual improvement in progressive motility
# 3.15 percentage points better motility
delta_motility <- 3.15

data_app_mot <- data_app_mot %>%
  mutate(
    motility_shifted = prog_motility_pct + delta_motility,
    motility_shifted = pmin(pmax(motility_shifted, 0), 100)
  )

# percentage below 32% after improved motility
shifted_p_below <- mean(data_app_mot$motility_shifted < 32)
shifted_p_below

# reduction in proportion below threshold
excess_p_below <- p0_motility_dk - shifted_p_below
excess_p_below
```

## Estimating population size

```{r}
# population size of men 15-49 men in DK
# https://www.statbank.dk/FOLK2
N_men_15_49 <- 1341884

# scaling to Denmark
# how many men in my selected age group will be affected by the shift in motility caused by BPS exposure?
AC_bu <- excess_p_below * N_men_15_49
```

## PLOTTING

```{r}


# create long format for plotting
data_app_plot <- data_app_mot %>%
  dplyr::select(
    Original = prog_motility_pct,
    After_shift = motility_shifted
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "scenario",
    values_to = "motility_pct"
  )

# density plot
ggplot(data_app_plot, aes(x = motility_pct, fill = scenario)) +
  geom_density(alpha = 0.35) +
  geom_vline(xintercept = 32, linetype = "dashed", linewidth = 1) +
  labs(
    x = "Progressive motility (%)",
    y = "Density",
    fill = "Scenario",
    title = "Applicant progressive motility before and after BPS-related shift",
    subtitle = "Dashed line indicates WHO 32% progressive motility threshold"
  ) +
  theme_classic()




p_below_summary <- data_app_mot %>%
  mutate(
    Original = prog_motility_pct,
    After_shift = motility_shifted
  ) %>%
  dplyr::select(Original, After_shift) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "scenario",
    values_to = "motility_pct"
  ) %>%
  group_by(scenario) %>%
  summarise(
    percent_below_32 = 100 * mean(motility_pct < 32),
    .groups = "drop"
  )

ggplot(p_below_summary, aes(x = scenario, y = percent_below_32, fill = scenario)) +
  geom_col(width = 0.6) +
  labs(
    x = NULL,
    y = "Applicants below 32% progressive motility (%)",
    title = "Proportion of applicants below WHO threshold before and after BPS-related shift"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
getwd()
write_xlsx(results_semen, "results_semen_china.xlsx")
```

### BOTTOM UP WITH DOSE-RESPONSE

```{r}
# Table 4: BPS and progressive motility
# Values are percentage point changes, not relative %
pct_q1 <- 0
pct_q2 <- 1.25
pct_q3 <- -0.86
pct_q4 <- -3.15

# Back-calculate SEs from 95% CIs
se_q2_cont <- (4.41 - (-1.90)) / (2 * 1.96)
se_q3_cont <- (4.01 - (-2.29)) / (2 * 1.96)
se_q4_cont <- (0.02 - (-6.32)) / (2 * 1.96)

study_dat_cont <- tibble(
  dose = c(dose_q1, dose_q2, dose_q3, dose_q4),
  dose_c = dose - dose_q1,
  motility_change_pp = c(pct_q1, pct_q2, pct_q3, pct_q4),
  se = c(NA, se_q2_cont, se_q3_cont, se_q4_cont)
)

fit_motility <- lm(
  motility_change_pp ~ 0 + dose_c,
  data = study_dat_cont[2:4, ],
  weights = 1 / se^2
)

beta_motility <- unname(coef(fit_motility)["dose_c"])
beta_motility_se <- summary(fit_motility)$coefficients["dose_c", "Std. Error"]

beta_motility
beta_motility_se
```

### Apply the dose-response to Danish exposure

```{r}
# Danish exposure level
median_dk <- df$median_dk
x_ref <- dose_q1

# Estimated BPS-related change in motility at Danish exposure
# This is in percentage points
# Danish BPS exposure is predicted to reduce progressive sperm motility by x percentage points
delta_motility <- beta_motility * (median_dk - x_ref)


delta_motility
```

### what would motility look like with no BPS exposure?

```{r}
data_app_mot_bu <- data_app_mot %>%
  mutate(
    observed_motility = prog_motility_pct,

    # If delta_motility is negative, removing BPS increases motility
    no_bps_motility = observed_motility - delta_motility,

    # keep values between 0 and 100
    no_bps_motility = pmin(pmax(no_bps_motility, 0), 100)
  )


# CALCULATING THE EFFECT

p_below_observed <- mean(data_app_mot_bu$observed_motility < 32)
p_below_no_bps   <- mean(data_app_mot_bu$no_bps_motility < 32)

excess_p_below <- p_below_observed - p_below_no_bps

p_below_observed
p_below_no_bps
excess_p_below

# SCALED TO DANISH MEN?

N_men_15_49 <- 1341884

AC_bu <- excess_p_below * N_men_15_49

AC_bu
```

### plotting :-)

```{r}
data_app_plot_bu <- data_app_mot_bu %>%
  dplyr::select(
    Observed = observed_motility,
    No_BPS_exposure = no_bps_motility
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "scenario",
    values_to = "motility_pct"
  )

ggplot(data_app_plot_bu, aes(x = motility_pct, fill = scenario)) +
  geom_density(alpha = 0.35) +
  geom_vline(xintercept = 32, linetype = "dashed", linewidth = 1) +
  labs(
    x = "Progressive motility (%)",
    y = "Density",
    fill = "Scenario",
    title = "Bottom-up dose-response estimate of BPS impact on progressive motility",
    subtitle = "Counterfactual assumes no BPS exposure; dashed line = WHO 32% threshold"
  ) +
  theme_classic()
```

```{r}
p_below_summary_bu <- data_app_plot_bu %>%
  group_by(scenario) %>%
  summarise(
    percent_below_32 = 100 * mean(motility_pct < 32),
    .groups = "drop"
  )

ggplot(p_below_summary_bu, aes(x = scenario, y = percent_below_32, fill = scenario)) +
  geom_col(width = 0.6) +
  labs(
    x = NULL,
    y = "Applicants below 32% progressive motility (%)",
    title = "Estimated proportion below WHO threshold with and without BPS exposure"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

## Normal distribution approach

```{r}
# Fit normal distribution to Danish motility data
mu_obs <- mean(data_app_mot$prog_motility_pct)
sd_obs <- sd(data_app_mot$prog_motility_pct)

# Dose-response estimated BPS effect at Danish exposure
delta_motility <- beta_motility * (median_dk - x_ref)

# Counterfactual mean without BPS exposure
# if delta is negative, this increases the mean
mu_no_bps <- mu_obs - delta_motility

# Proportion below WHO threshold
p_below_obs <- pnorm(32, mean = mu_obs, sd = sd_obs)
p_below_no_bps <- pnorm(32, mean = mu_no_bps, sd = sd_obs)

# Attributable proportion below threshold
excess_p_below <- p_below_obs - p_below_no_bps

p_below_obs
p_below_no_bps
excess_p_below

# scaled to Danish men?
AC_poor_motility_bu <- excess_p_below * N_men_15_49
AC_poor_motility_bu


# same mediation fraction as top-down
med_frac <- 0.40

# convert poor motility cases to infertility cases
AC_infertility_bu <- AC_poor_motility_bu * med_frac
AC_infertility_bu
```

```{r}
plot_grid_normal <- tibble(
  motility_pct = seq(0, 100, length.out = 500),
  Observed = dnorm(motility_pct, mean = mu_obs, sd = sd_obs),
  No_BPS_exposure = dnorm(motility_pct, mean = mu_no_bps, sd = sd_obs)
) %>%
  pivot_longer(
    cols = c(Observed, No_BPS_exposure),
    names_to = "scenario",
    values_to = "density"
  )

ggplot(plot_grid_normal, aes(x = motility_pct, y = density, fill = scenario)) +
  geom_area(alpha = 0.35, position = "identity") +
  geom_line(aes(color = scenario), linewidth = 1) +
  geom_vline(xintercept = 32, linetype = "dashed", linewidth = 1) +
  labs(
    x = "Progressive motility (%)",
    y = "Density",
    fill = "Scenario",
    color = "Scenario",
    title = "Bottom-up normal-distribution model of BPS impact on progressive motility",
    subtitle = "Counterfactual shifts the mean according to the BPS dose-response; dashed line = WHO 32% threshold"
  ) +
  theme_classic()
```

## infertility plot?

```{r}
infertility_summary <- tibble(
  outcome = c("Poor motility cases", "Infertility cases"),
  cases = c(AC_poor_motility_bu, AC_infertility_bu)
)

ggplot(infertility_summary, aes(x = outcome, y = cases, fill = outcome)) +
  geom_col(width = 0.6) +
  labs(
    x = NULL,
    y = "Attributable cases",
    title = "Estimated BPS-attributable cases from bottom-up model",
    subtitle = "Infertility cases estimated by applying mediation fraction to poor-motility cases"
  ) +
  theme_classic() +
  theme(legend.position = "none")





infertility_burden <- tibble(
  scenario = c("Observed", "No_BPS_exposure"),
  poor_motility_cases = c(p_below_obs, p_below_no_bps) * N_men_15_49,
  infertility_cases = poor_motility_cases * med_frac
)

ggplot(infertility_burden, aes(x = scenario, y = infertility_cases, fill = scenario)) +
  geom_col(width = 0.6) +
  labs(
    x = NULL,
    y = "Estimated infertility cases",
    title = "Estimated infertility cases with and without BPS exposure",
    subtitle = "Based on proportion below 32% progressive motility and mediation fraction"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r}
set.seed(123)

x_dk <- rlnorm(
  100000,
  meanlog = unique(df$meanlog_dk),
  sdlog = unique(df$sdlog_dk)
)

hist(x_dk, breaks = 100)

hist(x_dk[x_dk < 1], breaks = 100)
```
