---
title: "diabetes modeling korea spline"
format: html
editor: visual
---

## Libraries

```{r}
library(tidyverse)
library(readxl)
library(rriskDistributions)
library(writexl)
library(MASS)
```

## Importing data

```{r}
set.seed(123)

# loading danish inputs, sex, LOD, BPS percentiles & diabetes incidence
df <- read_excel("/Users/idasofieadrian/Desktop/DTU/SPECIALE/Speciale/data/diabetes korea (GBD).xlsx") %>%
  mutate(
    meanlog_dk = log(p_50),
    sdlog_dk = (log(p_75) - log(p_25)) / (qnorm(0.75) - qnorm(0.25)),

    # expected fraction of population above LOD
    p_detectable_dk = 1 - plnorm(LOD, meanlog_dk, sdlog_dk),

    # arithmetic mean exposure
    mean_exp_dk = exp(meanlog_dk + 0.5 * sdlog_dk^2),

    # median exposure
    median_dk = p_50
  )
```

## Korean study inputs, BPS and diabetes

```{r}
# Supplementary Table S6: DM ORs among participants with BPS > LOD
OR_t2  <- 1.73
LCL_t2 <- 1.00
UCL_t2 <- 2.98

OR_t3  <- 2.13
LCL_t3 <- 1.29
UCL_t3 <- 3.50

# Convert 95% CI to log-scale SE
se_t2 <- (log(UCL_t2) - log(LCL_t2)) / (2 * 1.96)
se_t3 <- (log(UCL_t3) - log(LCL_t3)) / (2 * 1.96)
```

## Korean detectable BPS distribution

```{r}
# Supplementary Table S1: detectable BPS distribution (> LOD)
fit_kor <- get.lnorm.par(
  p = c(0.25, 0.50, 0.75),
  q = c(0.029, 0.049, 0.104),
  show.output = FALSE,
  plot = TRUE
)

meanlog_kor <- unname(fit_kor["meanlog"])
sdlog_kor   <- unname(fit_kor["sdlog"])

# Simulate Korean detectable exposure distribution
x_kor <- rlnorm(1e6, meanlog = meanlog_kor, sdlog = sdlog_kor)

# Tertile cutpoints among detectable Korean values
cut1 <- qlnorm(1/3, meanlog = meanlog_kor, sdlog = sdlog_kor)
cut2 <- qlnorm(2/3, meanlog = meanlog_kor, sdlog = sdlog_kor)

# Median dose within each tertile
dose_t1 <- mean(x_kor[x_kor <= cut1])
dose_t2 <- mean(x_kor[x_kor > cut1 & x_kor <= cut2])
dose_t3 <- mean(x_kor[x_kor > cut2])
```

## Build dose-response dataset

```{r}
study_dat <- tibble(
  dose   = c(dose_t1, dose_t2, dose_t3),
  dose_c = c(dose_t1, dose_t2, dose_t3) - dose_t1,
  logrr  = log(c(1, OR_t2, OR_t3)),
  se     = c(NA, se_t2, se_t3)
)

study_dat
```

## Fit quadratic model

```{r}
# Give the reference point a very large weight so the curve stays anchored at (dose_c = 0, logrr = 0)
study_dat2 <- study_dat %>%
  mutate(w = c(1e6, 1 / se_t2^2, 1 / se_t3^2))

fit_quad <- lm(
  logrr ~ 0 + dose_c + I(dose_c^2),
  data = study_dat2,
  weights = w
)

summary(fit_quad)

coef_quad <- coef(fit_quad)
vcov_quad <- vcov(fit_quad)

coef_quad
vcov_quad
```

## Plot quadratic fit

```{r}
pred_grid_quad <- tibble(
  dose = seq(min(study_dat2$dose), max(study_dat2$dose), length.out = 100)
)

pred_grid_quad$dose_c <- pred_grid_quad$dose - dose_t1
pred_grid_quad$logrr_fit <- predict(fit_quad, newdata = pred_grid_quad)

p_logrr_quad <- ggplot() +
  geom_line(
    data = pred_grid_quad, 
    aes(x = dose, y = logrr_fit),
    color = "#045A8DFF") +
  geom_point(
    data = study_dat2, 
    aes(x = dose, y = logrr), 
    size = 3,
    color = "#045A8DFF") +
  labs(title = "BPS and diabetes risk (quadratic fit)",
       x = "Urinary BPS concentration",
       y = "log(OR)") +
  theme_classic() +
  theme(plot.title = element_text(size = 16, face = "bold"))


p_logrr_quad

ggsave(
  filename = "logrr_quadratic_model_korea.png",
  plot = p_logrr_quad,
  width = 7,
  height = 5,
  dpi = 300
)
```

## Inputs for YLD

```{r}
# Median age at onset of type 2 diabetes in DK
age_m <- 62.1
age_f <- 64.6

# Life expectancy in DK
life_m <- 78.8
life_f <- 82.8

# estimating duration
D_m <- life_m - age_m
D_f <- life_f - age_f

df <- df %>%
  mutate(
    D_years = case_when(
      str_detect(group, "^Male")   ~ D_m,
      str_detect(group, "^Female") ~ D_f,
      TRUE ~ NA_real_
    )
  )
```

## Inputs for YLL

```{r}
# deaths and YLL from GBD (DK + T2D)
deaths_f <- 566.89
deaths_m <- 715.34
yll_f    <- 7837.82
yll_m    <- 12702.54

# standard expected years of life lost
seyll_f <- yll_f / deaths_f
seyll_m <- yll_m / deaths_m

df <- df %>%
  mutate(
    deaths = case_when(
      str_detect(group, "^Female") ~ deaths_f,
      str_detect(group, "^Male")   ~ deaths_m,
      TRUE ~ NA_real_
    ),
    seyll = case_when(
      str_detect(group, "^Female") ~ seyll_f,
      str_detect(group, "^Male")   ~ seyll_m,
      TRUE ~ NA_real_
    )
  )
```

## Disability weight inputs

```{r}
# downloaded from GBD
YLD_total       <- 13225.904475796962 + 8355.60842682979
YLD_lower_total <-  8916.644865520693 + 5617.300593266269
YLD_upper_total <- 17957.837169143662 + 11465.14910195645

prev_total       <- 194510.88289338743 + 122399.73795153094
prev_lower_total <- 169049.8026957066  + 111088.08159363415
prev_upper_total <- 211790.44277738212 + 134597.21198640313

DW_mode <- YLD_total / prev_total
DW_min  <- YLD_lower_total / prev_upper_total
DW_max  <- YLD_upper_total / prev_lower_total

DW_mode
DW_min
DW_max
```

## Helper functions

```{r}
# PERT distribution for disability weight uncertainty
rpert <- function(n, min, mode, max) {
  alpha <- 1 + 4 * (mode - min) / (max - min)
  beta  <- 1 + 4 * (max - mode) / (max - min)
  min + rbeta(n, alpha, beta) * (max - min)
}
```

## Monte Carlo

```{r}
run_mc_row_daly_korea_quad <- function(
  i_diabetes,
  median_dk,
  p_detectable_dk,
  D_years,
  deaths,
  seyll,
  DW_min,
  DW_mode,
  DW_max,
  coef_quad,
  vcov_quad,
  dose_t1,
  baseline_uncertainty = FALSE
) {

  # draw one possible set of quadratic coefficients
  beta_draw <- MASS::mvrnorm(1, mu = coef_quad, Sigma = vcov_quad)

  b1 <- beta_draw[1]
  b2 <- beta_draw[2]

  # centered Danish exposure
  dose_c_dk <- median_dk - dose_t1

  # predicted log(RR) from quadratic model
  logrr_mean <- b1 * dose_c_dk + b2 * dose_c_dk^2

  # convert to RR
  rr_mean <- exp(logrr_mean)

  # population attributable fraction
  PAF <- (p_detectable_dk * (rr_mean - 1)) / (1 + p_detectable_dk * (rr_mean - 1))

  base_cases <- if (baseline_uncertainty) {
    rpois(1, lambda = i_diabetes)
  } else {
    i_diabetes
  }

  AC <- PAF * base_cases
  DW <- rpert(1, DW_min, DW_mode, DW_max)
  YLD <- AC * D_years * DW
  AD <- PAF * deaths
  YLL <- AD * seyll
  DALY <- YLD + YLL

  tibble(
    PAF = PAF,
    AC = AC,
    AD = AD,
    YLD = YLD,
    YLL = YLL,
    DALY = DALY
  )
}
```

## Monte Carlo summary function

```{r}
run_mc_summary_daly_korea_quad <- function(
  i_diabetes,
  median_dk,
  p_detectable_dk,
  D_years,
  deaths,
  seyll,
  DW_min,
  DW_mode,
  DW_max,
  coef_quad,
  vcov_quad,
  dose_t1,
  nsim_outer = 3000,
  baseline_uncertainty = FALSE
) {

  sims <- map_dfr(seq_len(nsim_outer), ~ run_mc_row_daly_korea_quad(
    i_diabetes = i_diabetes,
    median_dk = median_dk,
    p_detectable_dk = p_detectable_dk,
    D_years = D_years,
    deaths = deaths,
    seyll = seyll,
    DW_min = DW_min,
    DW_mode = DW_mode,
    DW_max = DW_max,
    coef_quad = coef_quad,
    vcov_quad = vcov_quad,
    dose_t1 = dose_t1,
    baseline_uncertainty = baseline_uncertainty
  ))

  tibble(
    PAF_med = median(sims$PAF, na.rm = TRUE),
    PAF_lo  = quantile(sims$PAF, 0.025, na.rm = TRUE),
    PAF_hi  = quantile(sims$PAF, 0.975, na.rm = TRUE),

    AC_med = median(sims$AC, na.rm = TRUE),
    AC_lo  = quantile(sims$AC, 0.025, na.rm = TRUE),
    AC_hi  = quantile(sims$AC, 0.975, na.rm = TRUE),

    AD_med = median(sims$AD, na.rm = TRUE),
    AD_lo  = quantile(sims$AD, 0.025, na.rm = TRUE),
    AD_hi  = quantile(sims$AD, 0.975, na.rm = TRUE),

    YLD_med = median(sims$YLD, na.rm = TRUE),
    YLD_lo  = quantile(sims$YLD, 0.025, na.rm = TRUE),
    YLD_hi  = quantile(sims$YLD, 0.975, na.rm = TRUE),

    YLL_med = median(sims$YLL, na.rm = TRUE),
    YLL_lo  = quantile(sims$YLL, 0.025, na.rm = TRUE),
    YLL_hi  = quantile(sims$YLL, 0.975, na.rm = TRUE),

    DALY_med = median(sims$DALY, na.rm = TRUE),
    DALY_lo  = quantile(sims$DALY, 0.025, na.rm = TRUE),
    DALY_hi  = quantile(sims$DALY, 0.975, na.rm = TRUE)
  )
}
```

## Run Monte Carlo

```{r}
results_daly_korea_quad <- df %>%
  mutate(
    mc = pmap(
      list(i_diabetes, median_dk, p_detectable_dk, D_years, deaths, seyll),
      ~ run_mc_summary_daly_korea_quad(
        i_diabetes = ..1,
        median_dk = ..2,
        p_detectable_dk = ..3,
        D_years = ..4,
        deaths = ..5,
        seyll = ..6,
        DW_min = DW_min,
        DW_mode = DW_mode,
        DW_max = DW_max,
        coef_quad = coef_quad,
        vcov_quad = vcov_quad,
        dose_t1 = dose_t1,
        nsim_outer = 3000,
        baseline_uncertainty = FALSE
      )
    )
  ) %>%
  unnest(mc)

results_daly_korea_quad
```

## plotting quadratic model

```{r}
pred_grid_quad <- tibble(
  dose = seq(min(study_dat2$dose), max(df$median_dk), length.out = 200)
) %>%
  mutate(dose_c = dose - dose_t1) %>%
  bind_cols(
    predict(fit_quad, newdata = ., interval = "confidence", level = 0.95) %>%
      as_tibble() %>%
      rename(logrr_fit = fit, logrr_lwr = lwr, logrr_upr = upr)
  )

study_dat2_plot <- study_dat2 %>%
  mutate(
    logrr_lower = logrr - 1.96 * se,
    logrr_upper = logrr + 1.96 * se
  )

p_quad <- ggplot() +
  geom_ribbon(
    data = pred_grid_quad,
    aes(x = dose, ymin = logrr_lwr, ymax = logrr_upr),
    fill = "#756bb1",
    alpha = 0.15
  ) +
  geom_line(
    data = pred_grid_quad,
    aes(x = dose, y = logrr_fit),
    color = "#756bb1",
    linewidth = 1
  ) +
  geom_errorbar(
    data = study_dat2_plot,
    aes(x = dose, ymin = logrr_lower, ymax = logrr_upper),
    color = "#54278f",
    width = 0.006,
    linewidth = 0.6
  ) +
  geom_point(
    data = study_dat2_plot,
    aes(x = dose, y = logrr),
    color = "#54278f",
    size = 3
  ) +
  geom_segment(
    aes(
      x = min(study_dat2$dose),
      xend = max(df$median_dk),
      y = 0,
      yend = 0
    ),
    linetype = "dashed",
    linewidth = 0.4,
    color = "grey40"
  ) +
  geom_vline(
    xintercept = unique(df$median_dk),
    linetype = "dashed"
  ) +
  labs(
    title = "Quadratic dose-response model for BPS and T2D risk",
    x = "Urinary BPS concentration",
    y = "log(OR)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 13)
  )

p_quad

# saving plot
ggsave(
  filename = "quad_model_korea.png",
  plot = p_quad,
  width = 7,
  height = 5,
  dpi = 300
)
```

## plotting spline model

```{r}
library(splines)

fit_spline <- lm(
  logrr ~ 0 + ns(dose_c, df = 2),
  data = study_dat2,
  weights = w
)

summary(fit_spline)

# making a sequence of dose values for plotting
# now also adding model uncertainty 
pred_grid_spline <- tibble(
  dose = seq(min(study_dat2$dose), max(df$median_dk), length.out = 100)
) %>%
  mutate(dose_c = dose - dose_t1) %>%
  bind_cols(
    predict(fit_spline, newdata = ., interval = "confidence", level = 0.95) %>%
      as_tibble() %>%
      rename(logrr_fit = fit, logrr_lwr = lwr, logrr_upr = upr)
  )
study_dat2_plot <- study_dat2 %>%
  mutate(
    logrr_lower = logrr - 1.96 * se,
    logrr_upper = logrr + 1.96 * se
  )

p_spline <- ggplot() +
  geom_ribbon(
    data = pred_grid_spline,
    aes(x = dose, ymin = logrr_lwr, ymax = logrr_upr),
    fill = "#756bb1",
    alpha = 0.15
  ) +
  geom_line(
    data = pred_grid_spline,
    aes(x = dose, y = logrr_fit),
    color = "#756bb1",
    linewidth = 1
  ) +
  geom_errorbar(
    data = study_dat2_plot,
    aes(x = dose, ymin = logrr_lower, ymax = logrr_upper),
    color = "#54278f",
    width = 0.006,
    linewidth = 0.6
  ) +
  geom_point(
    data = study_dat2_plot,
    aes(x = dose, y = logrr),
    color = "#54278f",
    size = 3
  ) +
  geom_segment(
    aes(
      x = min(study_dat2$dose),
      xend = max(study_dat2$dose),
      y = 0,
      yend = 0
    ),
    linetype = "dashed",
    linewidth = 0.4,
    color = "grey40"
  ) +
  geom_vline(
    xintercept = unique(df$median_dk),
    linetype = "dashed"
  ) +
  labs(
    title = "Spline dose-response model for BPS and T2D risk",
    x = "Urinary BPS concentration",
    y = "log(OR)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 13)
  )

p_spline

# saving plot
ggsave(
  filename = "spline_model_korea.png",
  plot = p_spline,
  width = 7,
  height = 5,
  dpi = 300
)
```

## Monotonocity?

```{r}
# mono
library(scam)

study_dat2 <- study_dat2 %>%
  mutate(
    dose_c_scaled = dose_c / max(dose_c)
  )

fit_mono_spline <- scam(
  logrr ~ 0 + s(dose_c_scaled, bs = "mpi", k = 4),
  data = study_dat2,
  weights = w
)

summary(fit_mono_spline)


dose_scale <- max(study_dat2$dose_c, na.rm = TRUE)

pred_grid_mono <- tibble(
  dose = seq(
    min(study_dat2$dose),
    max(df$median_dk),
    length.out = 200
  )
) %>%
  mutate(
    dose_c = dose - dose_t1,
    dose_c_scaled = dose_c / dose_scale
  )

pred_grid_mono <- pred_grid_mono %>%
  mutate(
    logrr_fit = predict(fit_mono_spline, newdata = pred_grid_mono),
    rr_fit = exp(logrr_fit)
  )

p_mono_spline <- ggplot(pred_grid_mono, aes(x = dose, y = rr_fit)) +
  geom_line(color = "#756bb1", linewidth = 1) +
  geom_point(
    data = study_dat2,
    aes(x = dose, y = exp(logrr)),
    inherit.aes = FALSE,
    size = 3,
    color = "#54278f"
  ) +
  geom_vline(
    xintercept = unique(df$median_dk),
    linetype = "dashed"
  ) +
  labs(
    title = "Monotone spline exposure-response model",
    x = "Urinary BPS concentration",
    y = "RR"
  ) +
  theme_minimal()

p_mono_spline


#No really working...
```

```{r}
xmax <- max(df$median_dk, study_dat2$dose, na.rm = TRUE) - dose_t1

x_constraint <- seq(0, xmax, length.out = 200)

objective_quad <- function(beta) {
  b1 <- beta[1]
  b2 <- beta[2]

  pred <- b1 * study_dat2$dose_c + b2 * study_dat2$dose_c^2

  sum(study_dat2$w * (study_dat2$logrr - pred)^2)
}

ui <- cbind(
  1,
  2 * x_constraint
)

ci <- rep(0, length(x_constraint))

fit_mono_quad <- constrOptim(
  theta = coef(fit_quad),
  f = objective_quad,
  grad = NULL,
  ui = ui,
  ci = ci
)

coef_mono_quad <- fit_mono_quad$par
coef_mono_quad


pred_grid_mono_quad <- tibble(
  dose = seq(
    min(study_dat2$dose),
    max(df$median_dk, study_dat2$dose, na.rm = TRUE),
    length.out = 200
  )
) %>%
  mutate(
    dose_c = dose - dose_t1,
    logrr_fit = coef_mono_quad[1] * dose_c + coef_mono_quad[2] * dose_c^2,
    rr_fit = exp(logrr_fit)
  )

p_mono_quad <- ggplot(pred_grid_mono_quad, aes(x = dose, y = rr_fit)) +
  geom_line(color = "#045A8DFF", linewidth = 1) +
  geom_point(
    data = study_dat2,
    aes(x = dose, y = exp(logrr)),
    inherit.aes = FALSE,
    size = 3,
    color = "#045A8DFF"
  ) +
  geom_vline(
    xintercept = unique(df$median_dk),
    linetype = "dashed"
  ) +
  labs(
    title = "Monotone quadratic exposure-response model",
    x = "Urinary BPS concentration",
    y = "RR"
  ) +
  theme_classic() +
  theme(plot.title = element_text(size = 16, face = "bold"))

p_mono_quad
```
